
```{r}
library(tidyverse)
library(lubridate)
library(modelr)
```

#Read in data & omit NA functions
```{r}
people <- read.csv("C:/Users/pgiom/Downloads/R/People_CSV.csv") %>%
  mutate(
    dob = dmy(dob),
    debut = ymd(debut),
    finalGame = ymd(finalGame),
    Month = month(dob, label = TRUE),
    finalYear = year(finalGame),
    debutYear = year(debut),
    careerYears = finalYear - debutYear
  ) %>%
  select(careerYears, weight, height) %>%
  na.omit()

batting <- read.csv("C:/Users/pgiom/Downloads/R/Batting.csv")
pitching <- read.csv("C:/Users/pgiom/Downloads/Pitching.csv")
salaries <- read.csv("C:/Users/pgiom/Downloads/Salaries.csv")
people <- read.csv("C:/Users/pgiom/Downloads/R/People_CSV.csv")
```

#Seasons by weight

```{r}
seasonsByWeight <- people %>%
  group_by(careerYears) %>%
  summarize(averageWeight = mean(weight), averageHeight = mean(height), .groups = "drop")
```

#Plot of Seasons by Weight
```{r}
ggplot(seasonsByWeight, aes(x = careerYears, y = averageWeight)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(title = "Average Weight by Career Years", x = "Career Years", y = "Average Weight")
```


# Regression model: Store the model object for later use and avoid redundant computations.
```{r}
linearRegModel <- lm(averageWeight ~ careerYears, data = seasonsByWeight)
summary(linearRegModel)
```


# Partition the data into test and training sets using set.seed for reproducibility.
```{r}
set.seed(123) # Set seed for reproducibility
partition <- sort(sample(nrow(seasonsByWeight), nrow(seasonsByWeight)*.8))
train_set <- seasonsByWeight[partition, ]
test_indices <- setdiff(1:nrow(seasonsByWeight), partition)
test_set <- seasonsByWeight[test_indices, ]

# Fit the model using the training set
linRegTrain <- lm(averageWeight ~ careerYears, data = train_set)
summary(linRegTrain)

# Make predictions and calculate residuals on the test set
test_set %>%
  add_predictions(linRegTrain) %>%
  add_residuals(linRegTrain) -> testSet

# Calculate the average of residuals
avg_resid <- mean(testSet$resid)
avg_resid

# Calculate and summarize the mean squared error (MSE)
testSet %>%
  mutate(sq_error = resid^2) %>%
  summarize(MSE = mean(sq_error)) -> testSetSummary

testSetSummary$MSE


```


# Use the model to predict on the training set, then calculate MSE on the test set.
```{r}
linRegTrain <- lm(averageWeight ~ careerYears, data = train_set)
summary(linRegTrain)

predictions <- predict(linRegTrain, newdata = test_set)
MSE <- mean((test_set$averageWeight - predictions) ^ 2)
MSE
```
#Plots
```{r}
# Plot Runs vs StrikeOuts for pitchers
ggplot(data = pitching, aes(x = SO, y =R )) +
  geom_point(alpha = 0.5) +  # setting alpha for better visibility if points overlap
  labs(title = "Runs Allowed vs. Strikeouts for Pitchers",
       x = "Runs Allowed",
       y = "Strikeouts") +
  theme_minimal()
```
#Salaries
```{r}
#Top Hitter / Top Pitcher
best_hitter <- batting[which.max(batting$HR), ]

# Filter out the data for the two best players
players_salaries <- salaries %>%
  filter(playerID %in% c("bondsba01", "johnsra05"))

#bar chart for salaries
ggplot(players_salaries, aes(x = factor(yearID), y = salary, fill = playerID)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Salary", fill = "Player ID") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format()) +
  ggtitle("Salary Comparison of bondsba01 and johnsra05") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#BAOpp vs. Strikeouts visualization
```{r}
#get rid of NA values and values greater than 0.5
clean_pitching <- pitching %>%
  na.omit() %>%
  filter(BAOpp <= 0.5) 
  
ggplot(clean_pitching, aes(x = BAOpp, y = SO, color=yearID)) +
  geom_point() +  # Use geom_point for scatter plot
  labs(x = "Opponent's Batting Average (BAOpp)", y = "Strike Outs (SO)",
       title = "Relationship between BAOpp and Strike Outs") +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE)  # Adds a linear regression line without confidence interval
```
#Runs vs. Strikeouts
```{r}
hist(pitching$ERA)
clean_pitching <- pitching %>%
  na.omit() %>%
  filter(ERA <=10)
  
#scatter plot with data mined 
ggplot(clean_pitching, aes(x = SO, y = ERA)) +
  geom_point() +  # Use geom_point for a scatter plot
  labs(x = "Strikeouts (SO)", y = "Earned Run Average (ERA)",
       title = "Relationship between Runs and Strikeouts for Pitchers") +
  theme_minimal() +
  geom_smooth()  
```

